Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2023-06-28 manuscript.Rmd
Updates

\Revision{A pattern where a general trend is not representing the individual trend is} shown in Panel A of Figure \@ref(fig:figure1) where some of the individual trends (i.e., slopes of the grey lines) differ from the general trend (i.e., slope of the blue line). If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. \Revision{Therefore, analyses on aggregated data do not appropriately account for differential experiment effects} [@van2021bayes; @van2023bayes]. As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder & Lu, 2005).

@@

\Revision{It is clear that neither of the alternative approaches yield satisfactory results by themselves. Conceptually, a hierarchical analysis provides a compromise between the aggregated and independent approach by drawing benefits from each method. The motivation for hierarchical models, as well as their general mathematical structure, has been covered by several tutorial articles} [e.g., @singmann2019introduction; @west2022linear]\Revision{, as well as text books (e.g., XX, XX).}

@@

Throughout the section, we assume that each observation $Y_{ijk}$ is the response time from participant $i$ in trial $k$ of condition $j$. The condition $j$ is the digit presented to the participant, and can take the values 2, 3, 4, 6, 7, and 8. We further assume that the digit side ($x$) is recorded using effect coding where a value of -0.5 indicates that the digit is smaller than 5, and a value of 0.5 indicates that the digit is larger than 5. Additionally, we assume that the digits presented in the trial are dummy-coded in the data using four variables ($u$, $v$, $w$, $z$). If digits 2 and 8 were presented all four variables are zero. The dummy variables represent the digits 3 ($u$), 4 ($v$), 6 ($w$), and 7 ($z$), and each take the value 1 if the respective digit was presented in the trial, and 0 if it was not.

@@

\begin{equation}
Y_{ijk} \sim \text{LogNormal}(\gamma_{i} + x_j \beta_{i} + u_j \delta_{7i} + v_j \delta_{6i} + w_j \delta_{4i} + z_j \delta_{3i}, \sigma^2).  
\end{equation}
Note that the only change in the model formulation is the change from a normal distribution to a log-normal distribution. \Revision{The density function describes the probability model that is assumed to be underlying the data. When the parameters of this probability model are known, the joint probability of the data can be inferred from the \Revision{density} function. The log-normal distribution only assigns probability to positive data values, and is, just as the RT distributions, right-skewed. Therefore, the log-normal distribution is a better match for the RT data.} The location parameter of the log-normal distribution is described through the same combination of parameters as the mean of the normal model. However, it has a different interpretation. For example, for a location parameter of $\mu = 0$ and a scale parameter of $\sigma^2 = 1$, the mode of the log-normal distribution is $0.368$, and the mean is $1.65$.

>>>>

\Revision{A pattern where the general trend does not reflect individual trends is shown in Panel A of Figure} \@ref(fig:figure1)\Revision{. The slopes of the grey lines reflect individual trends, and some of them differ considerably from the general trend shown as the slope of the blue line.} If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. \Revision{Therefore, analyses on aggregated data do not appropriately account for differential experiment effects} [@van2021bayes; @van2023bayes]. As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder & Lu, 2005).

@@

\Revision{It is clear that neither of the alternative approaches yield satisfactory results by themselves. Conceptually, a hierarchical analysis provides a compromise between the aggregated and independent approach by drawing benefits from each method. The motivation for hierarchical models, as well as their general mathematical structure, has been covered by several (tutorial) articles} [e.g., @efron1977stein; @singmann2019introduction; @west2022linear]\Revision{, as well as text books} [e.g., @gelman2006data; @hox2017multilevel; @nicenboim2023introduction].

@@

Throughout the section, we assume that each observation $Y_{ijk}$ is the response time from participant $i$ in trial $k$ of condition $j$. The condition $j$ is the digit presented to the participant, and can take the values 2, 3, 4, 6, 7, and 8. We further assume that the digit side ($x$) is recorded using effect coding where a value of -0.5 indicates that the digit is smaller than 5, and a value of 0.5 indicates that the digit is larger than 5. Additionally, we assume that the digits presented in the trial are dummy-coded in the data using four variables ($u$, $v$, $w$, $z$). If digits 2 and 8 were presented all four variables are zero. The dummy variables represent the digits 3 ($u$), 4 ($v$), 6 ($w$), and 7 ($z$), and each take the value 1 if the respective digit was presented in the trial, and 0 if it was not [for more on contrast coding, we refer to @pinheiro2006mixed; and @singmann2019introduction].

@@

\begin{equation}
Y_{ijk} \sim \text{LogNormal}(\gamma_{i} + x_j \beta_{i} + u_j \delta_{7i} + v_j \delta_{6i} + w_j \delta_{4i} + z_j \delta_{3i}, \sigma^2).  
\end{equation}

\Revision{Note that the only change in the model formulation is the change from a normal distribution to a log-normal distribution. The density function describes the probability model that is assumed to be underlying the data. When the parameters of this probability model are known, the joint probability of the data can be inferred from the density function. The log-normal distribution only assigns probability to positive data values, and is, just as the RT distributions, right-skewed. Therefore, the log-normal distribution is a better match for the RT data.} The location parameter of the log-normal distribution is described through the same combination of parameters as the mean of the normal model. However, it has a different interpretation. For example, for a location parameter of $\mu = 0$ and a scale parameter of $\sigma^2 = 1$, the mode of the log-normal distribution is $0.368$, and the mean is $1.65$.

### 2023-06-28 manuscript.tex
Updates

Our tutorial is directed at researchers who have a basic \textcolor{blue}{conceptual} understanding of Bayesian inference and parameter estimation and are planning to perform more complex Bayesian analyses \textcolor{blue}{of psychological phenomena. We hope to achieve a balance of detail and overview---while we want to provide guidance for a full Bayesian workflow, similar to Gelman et al.} (2020) \textcolor{blue}{and Schad et al.}(2021; 2022)\textcolor{blue}{, we don't shy away from in-depth explanations for the most curious scientists. To ensure that a broad audience is able to follow the tutorial, we limit and carefully explain equations. However, rudimentary knowledge of calculus is advantageous to fully understand Bayesian analysis. }

@@

\textcolor{blue}{A pattern where a general trend is not representing the individual trend is} shown in Panel A of Figure \ref{fig:figure1} where some of the individual trends (i.e., slopes of the grey lines) differ from the general trend (i.e., slope of the blue line). If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. \textcolor{blue}{Therefore, analyses on aggregated data do not appropriately account for the interdependence between observations} (Doorn, Aust, Haaf, Stefan, \& Wagenmakers, 2021, 2023). As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder \& Lu, 2005).

@@

\textcolor{blue}{It is clear that neither of the alternative approaches yield satisfactory results by themselves. Conceptually, a hierarchical analysis provides a compromise between the aggregated and independent approach by drawing benefits from each method. The motivation for hierarchical models, as well as their general mathematical structure, has been covered by several tutorial articles} (e.g., Singmann \& Kellen, 2019; West, Welch, \& Galecki, 2022)\textcolor{blue}{, as well as text books (e.g., XX, XX).}

\textcolor{blue}{In hierarchical analyses, the general or overall effect is sometimes called fixed effect, and the individual-specific deviations from this effect are in that case referred to as random effects (Rouder \& Lu, 2005). We can also estimate individual effects, not as deviations from an overarching effect, but as effects for each individual, as obtained from a per-participant analysis. Throughout this tutorial, we will use the terms that suit the interpretation of our analyses, namely general effect, individual deviation, and individual effect.}

@@

Throughout the section, we assume that each observation \(Y_{ijk}\) is the response time from participant \(i\) in trial \(k\) of condition \(j\). The condition \(j\) is the digit presented to the participant, and can take the values 2, 3, 4, 6, 7, and 8. We further assume that the digit side (\(x\)) is recorded using effect coding where a value of -0.5 indicates that the digit is smaller than 5, and a value of 0.5 indicates that the digit is larger than 5. Additionally, we assume that the digits presented in the trial are dummy-coded in the data using four variables (\(u\), \(v\), \(w\), \(z\)). If digits 2 and 8 were presented all four variables are zero. The dummy variables represent the digits 3 (\(u\)), 4 (\(v\)), 6 (\(w\)), and 7 (\(z\)), and each take the value 1 if the respective digit was presented in the trial, and 0 if it was not.

@@

\begin{equation}
Y_{ijk} \sim \text{LogNormal}(\gamma_{i} + x_j \beta_{i} + u_j \delta_{7i} + v_j \delta_{6i} + w_j \delta_{4i} + z_j \delta_{3i}, \sigma^2).  
\end{equation}
Note that the only change in the model formulation is the change from a normal distribution to a log-normal distribution. \textcolor{blue}{The density function describes the probability model that is assumed to be underlying the data. When the parameters of this probability model are known, the joint probability of the data can be inferred from the \textcolor{blue}{density} function. The log-normal distribution only assigns probability to positive data values, and is, just as the RT distributions, right-skewed. Therefore, the log-normal distribution is a better match for the RT data.} The location parameter of the log-normal distribution is described through the same combination of parameters as the mean of the normal model. However, it has a different interpretation. For example, for a location parameter of \(\mu = 0\) and a scale parameter of \(\sigma^2 = 1\), the mode of the log-normal distribution is \(0.368\), and the mean is \(1.65\).

@@

Another important issue is that wide priors are often \textcolor{blue}{improper probability distributions} (Hobert \& Casella, 1996). \textcolor{blue}{A prior is proper if two conditions hold: 1. All values of the probability distribution are positive definite, and 2. the probability distribution sums to one for discrete data or integrates to one for continuous data. An example of a proper and improper probability distribution are provided in Figure} \ref{fig:properfigure}. \textcolor{blue}{The figure shows two hypothetical priors for parameter $\theta$. The first condition can be assessed by checking the y-values---both priors have only positive definite values as function output (0.1, 0.2, and 0.4). To assess the second condition, we need to sum up the function values. Panel A shows a proper distribution where the values sum to 1; panel B shows an improper distribution where the values sum to 1.2.}

@@

For the symbolic distance effect, the estimated number of effective samples is provided in panel D of Figure \ref{fig:traceplots3p}. For the package \texttt{rstan} the number is generally high, much higher than the actual number of samples depicted by the green dashed line. The \texttt{brms} package shows much more variation. The lowest effective sample sizes can be found for the intercept parameters. Table 1 also displays the number of effective samples for the general effects parameters in the \texttt{rstan} model. If the number of effective samples is too low, \texttt{brms} and \texttt{rstan} will return a warning. Here, this was not case, and all other diagnostics were good, so we continue with the interpretation of the estimates. \textcolor{blue}{The difference between the model diagnostics of} \texttt{rstan} \textcolor{blue}{and} \texttt{brms} \textcolor{blue}{can be partly explained by the difference in model parametrization, such as the difference in the prior specification (i.e., specifying priors on variances versus priors on standard deviations). Although we tried to make the models as similar as possible, they are not identical.}

@@

Imagine this results in \(\text{BF}_{\text{nl}} = 100\). This means that the data are 100 times more likely under the normal model than under the log-normal model, clear evidence in favor of the normal model. \textcolor{blue}{But how do we know whether a Bayes factor of 100-to-1, or 10-to-1 or 3-to-1 fulfills?} As a reminder, Bayes factors are the relative predictive accuracy of two models for the observed data. Therefore, Bayes factors are odds or ratios. Odds themselves are directly interpretable without the need for decisions or cutoffs. For example, if a presidential candidate is favored 10-to-1 over another one, then these are just the odds, and it is beside the point whether these are large or not. According to this line of thought, Rouder et al. (2018) argue that researchers should take a similar approach when interpreting and reporting Bayes factors. While there exist rules of thumb to categorize Bayes factors verbally, we suggest researchers use their research context to form substantive conclusions based on Bayes factors, not arbitrary rules of thumb.

@@

First, we kept the discussion of the software packages and code fairly brief. We provide further code examples for the digit classification task in the \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{Online Supplement}. For different models, it may be useful to study additional tutorials and books. For \texttt{rstan}, we recommend the \texttt{Stan} manual (Stan Development Team, 2018b, 2018a). For \texttt{brms} there are several useful tutorials (Bürkner, 2017, 2018; Bürkner \& Vuorre, 2019). There are also several resources for Bayesian estimation with other samplers (Lee \& Wagenmakers, 2013; Rouder, Morey, \& Pratte, 2013). Two other useful tutorials for Bayesian modeling are Schad, Betancourt, and
Vasishth (2021) and Schad, Nicenboim, Bürkner, Betancourt, and Vasishth (2022). \textcolor{blue}{There are also other R software packages for Bayesian multilevel modeling besides} \texttt{brms} \textcolor{blue}{and} \texttt{rstan}\textcolor{blue}{, such as} \texttt{BayesFactor}(Morey \& Rouder, 2018) \textcolor{blue}{and} \texttt{rstanarm}(Goodrich, Gabry, Ali, \& Brilleman, 2020)\textcolor{blue}{, that we did not illustrate in this tutorial. However, most of these other R software packages are a bit more limited, for instance, in model specification and prior settings.}

@@

@@

@@

>>>>

Our tutorial is directed at researchers who have a basic \textcolor{blue}{conceptual} understanding of Bayesian inference and parameter estimation and are planning to perform more complex Bayesian analyses \textcolor{blue}{of psychological phenomena. We hope to achieve a balance of detail and overview---while we want to provide guidance for a full Bayesian workflow, similar to Gelman et al.} (2020) \textcolor{blue}{and Schad et al.}(2021; 2022)\textcolor{blue}{, we don't shy away from in-depth explanations for the most curious scientists. To ensure that a broad audience is able to follow the tutorial, we limit and carefully explain equations. However, rudimentary knowledge of calculus is advantageous to fully understand the Bayesian analyses. }

@@

\textcolor{blue}{A pattern where the general trend does not reflect individual trends is shown in Panel A of Figure} \ref{fig:figure1}\textcolor{blue}{. The slopes of the grey lines reflect individual trends, and some of them differ considerably from the general trend shown as the slope of the blue line.} If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. \textcolor{blue}{Therefore, analyses on aggregated data do not appropriately account for differential experiment effects} (Doorn, Aust, Haaf, Stefan, \& Wagenmakers, 2021, 2023). As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder \& Lu, 2005).

@@

\textcolor{blue}{It is clear that neither of the alternative approaches yield satisfactory results by themselves. Conceptually, a hierarchical analysis provides a compromise between the aggregated and independent approach by drawing benefits from each method. The motivation for hierarchical models, as well as their general mathematical structure, has been covered by several tutorial articles} (e.g., Singmann \& Kellen, 2019; West, Welch, \& Galecki, 2022)\textcolor{blue}{, as well as text books} (e.g., Gelman \& Hill, 2006; Hox, Moerbeek, \& Van de Schoot, 2017; Nicenboim, Schad, \& Vasishth, 2023).

\textcolor{blue}{In hierarchical analyses, the general or overall effect is sometimes called fixed effect, and the individual-specific deviations from this effect are in that case referred to as random effects (Rouder \& Lu, 2005). We can also estimate individual effects, not as deviations from an overarching effect, but as effects for each individual, as obtained from a per-participant analysis. Throughout this tutorial, we will use the terms general effect, individual deviation, and individual effect.}

@@

Throughout the section, we assume that each observation \(Y_{ijk}\) is the response time from participant \(i\) in trial \(k\) of condition \(j\). The condition \(j\) is the digit presented to the participant, and can take the values 2, 3, 4, 6, 7, and 8. We further assume that the digit side (\(x\)) is recorded using effect coding where a value of -0.5 indicates that the digit is smaller than 5, and a value of 0.5 indicates that the digit is larger than 5. Additionally, we assume that the digits presented in the trial are dummy-coded in the data using four variables (\(u\), \(v\), \(w\), \(z\)). If digits 2 and 8 were presented all four variables are zero. The dummy variables represent the digits 3 (\(u\)), 4 (\(v\)), 6 (\(w\)), and 7 (\(z\)), and each take the value 1 if the respective digit was presented in the trial, and 0 if it was not (for more on contrast coding, we refer to Pinheiro \& Bates, 2006; and Singmann \& Kellen, 2019).

@@

\begin{equation}
Y_{ijk} \sim \text{LogNormal}(\gamma_{i} + x_j \beta_{i} + u_j \delta_{7i} + v_j \delta_{6i} + w_j \delta_{4i} + z_j \delta_{3i}, \sigma^2).  
\end{equation}

\textcolor{blue}{Note that the only change in the model formulation is the change from a normal distribution to a log-normal distribution. The density function describes the probability model that is assumed to be underlying the data. When the parameters of this probability model are known, the joint probability of the data can be inferred from the density function. The log-normal distribution only assigns probability to positive data values, and is, just as the RT distributions, right-skewed. Therefore, the log-normal distribution is a better match for the RT data.} The location parameter of the log-normal distribution is described through the same combination of parameters as the mean of the normal model. However, it has a different interpretation. For example, for a location parameter of \(\mu = 0\) and a scale parameter of \(\sigma^2 = 1\), the mode of the log-normal distribution is \(0.368\), and the mean is \(1.65\).

@@

Another important issue is that wide priors are often \textcolor{blue}{improper probability distributions} (Hobert \& Casella, 1996). \textcolor{blue}{A prior is proper if two conditions hold: 1. All values of the probability distribution are equal or greater than zero, and 2. the probability distribution sums to one for discrete data or integrates to one for continuous data. An example of a proper and improper probability distribution are provided in Figure} \ref{fig:properfigure}. \textcolor{blue}{The figure shows two hypothetical priors for parameter $\theta$. The first condition can be assessed by checking the y-values---both priors have only positive values as function output (0.1, 0.2, and 0.4). To assess the second condition, we need to sum up the function values. Panel A shows a proper distribution where the values sum to 1; panel B shows an improper distribution where the values sum to 1.2.}

@@

For the symbolic distance effect, the estimated number of effective samples is provided in panel D of Figure \ref{fig:traceplots3p}. For the package \texttt{rstan} the number is generally high, much higher than the actual number of samples depicted by the green dashed line. The \texttt{brms} package shows much more variation. The lowest effective sample sizes can be found for the intercept parameters. Table 1 also displays the number of effective samples for the general effects parameters in the \texttt{rstan} model. If the number of effective samples is too low, \texttt{brms} and \texttt{rstan} will return a warning. Here, this was not case, and all other diagnostics were good, so we continue with the interpretation of the estimates. \textcolor{blue}{The difference between the model diagnostics of} \texttt{rstan} \textcolor{blue}{and} \texttt{brms} \textcolor{blue}{can be partly explained by the difference in model parametrization, such as the difference in the prior specification (i.e., specifying priors on variances versus priors on standard deviations). We equated the models where sensible which resulted in models that are fairly comparable. }

@@

Imagine this results in \(\text{BF}_{\text{nl}} = 100\). This means that the data are 100 times more likely under the normal model than under the log-normal model, clear evidence in favor of the normal model. \textcolor{blue}{But how do we know whether a Bayes factor of 100-to-1, or 10-to-1 or 3-to-1 is big enough?} As a reminder, Bayes factors are the relative predictive accuracy of two models for the observed data. Therefore, Bayes factors are odds or ratios. Odds themselves are directly interpretable without the need for decisions or cutoffs. For example, if a presidential candidate is favored 10-to-1 over another one, then these are just the odds, and it is beside the point whether these are large or not. According to this line of thought, Rouder et al. (2018) argue that researchers should take a similar approach when interpreting and reporting Bayes factors. While there exist rules of thumb to categorize Bayes factors verbally, we suggest researchers use their research context to form substantive conclusions based on Bayes factors, not arbitrary rules of thumb.

@@

First, we kept the discussion of the software packages and code fairly brief. We provide further code examples for the digit classification task in the \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{Online Supplement}. For different models, it may be useful to study additional tutorials and books. For \texttt{rstan}, we recommend the \texttt{Stan} manual (Stan Development Team, 2018b, 2018a). For \texttt{brms} there are several useful tutorials (Bürkner, 2017, 2018; Bürkner \& Vuorre, 2019). There are also several resources for Bayesian estimation with other samplers (Lee \& Wagenmakers, 2013; Rouder, Morey, \& Pratte, 2013). Two other useful tutorials for Bayesian modeling are Schad, Betancourt, and
Vasishth (2021) and Schad, Nicenboim, Bürkner, Betancourt, and Vasishth (2022). \textcolor{blue}{There are also other R software packages for Bayesian multilevel modeling besides} \texttt{brms} \textcolor{blue}{and} \texttt{rstan}\textcolor{blue}{, such as} \texttt{BayesFactor}(Morey \& Rouder, 2018) \textcolor{blue}{and} \texttt{rstanarm}(Goodrich, Gabry, Ali, \& Brilleman, 2020)\textcolor{blue}{, that we did not illustrate in this tutorial. However, most of these other R software packages are more limited, for instance, in model specification and prior settings.}

@@

\leavevmode\hypertarget{ref-gelman2006data}{}%
Gelman, A., \& Hill, J. (2006). \emph{Data analysis using regression and multilevel/hierarchical models}. Cambridge university press.

@@

\leavevmode\hypertarget{ref-hox2017multilevel}{}%
Hox, J. J., Moerbeek, M., \& Van de Schoot, R. (2017). \emph{Multilevel analysis: Techniques and applications}. Routledge.

@@

\leavevmode\hypertarget{ref-nicenboim2023introduction}{}%
Nicenboim, B., Schad, D., \& Vasishth, S. (2023). \emph{An introduction to bayesian data analysis for cognitive science}. Chapman; Hall/CRC Statistics in the Social; Behavioral Sciences Series.

\leavevmode\hypertarget{ref-pinheiro2006mixed}{}%
Pinheiro, J., \& Bates, D. (2006). \emph{Mixed-effects models in s and s-PLUS}. Springer science \& business media.

### 2023-06-28 propriety_section.Rmd
Updates

\Revision{Rather than submerging us in technical details or providing a formal justification of the Bayesian approach, we provide an intuition of Bayesian model comparison. A formal justification of Bayes factors is provided by @jeffreys1961theory, @kass1995bayes and @rouder2018teaching. The Bayes factor (BF) is a measure of how well one model predicts the data compared to another.}

>>>>

\Revision{Rather than submerging us in technical details or providing a formal justification of the Bayesian approach, we provide an intuition of Bayesian model comparison. A formal justification of Bayes factors is provided by} @jeffreys1961theory, @kass1995bayes and @rouder2018teaching. \Revision{The Bayes factor (BF) is a measure of how well one model predicts the data compared to another.}
